A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-20 FMCAD2021/1_intro.tex
some grammar mistakes

However, research progress of string constraint solving has been hampered by many major obstacles in both theory and tool implementation aspects (including long-standing open problems). 
Logical theories over strings have to allow string concatenation, which is arguably the most fundamental operation of strings. 
The most celebrated result concerning theories of strings is Makanin’s result on deciding the satisfiability problem for \emph{word equations}~\cite{makanin77}.
A simple example of a word equation is $xaby = ybax$, where $x, y$ are variables, and $a, b \in \Sigma$ are constant letters. 
A word equation is satisfiable if it has a solution, i.e., an assignment that maps variables to strings over the alphabet $\Sigma$ which equate the left-hand side with the right-hand side of the equation.
The correctness proof of Makanin’s algorithm is arguably one of the most complex termination proofs in computer science. 
Makanin’s result can be extended to include \emph{regular constraints} (a.k.a. regular expression matching, e.g., $x \in (ba)^*)$, and arbitrary Boolean connectives.
This extension is called word equations with regular constraints. 
However, the satisfiability problem of word equation together with length constraints (e.g., $|x|=|y|+1 \wedge wx=yx$) is still open.
The complexity of the satisfiability of word equations with regular constraints was proven to be PSPACE-complete by Plandowski~\cite{plandowski99}, after decades of improvement of the original algorithm by Makanin.

@@

The satisfiability of \emph{string/number conversion constraints}, e.g., $wx=yx \wedge |x| > \parseInt(y)$, has been proven undecidable in~\cite{DayGHMN18}. 
However, this kind of constraint is pervasive in scripting language programs. 
For example, it is common that programs read string input from text files and conversion a part of the string input to numbers.
Even more crucially, in many programming languages, the string-number conversion is a part of the definition of their core semantics. 
JavaScript, which powers most interactive content on the Web and increasingly server- side code with Node.js, is one of such languages.

@@

One more recent approach is flattening~\cite{Parosh:20:PLDI,Parosh:20:PLDI,AbdullaACDHRR18,AbdullaACDHRR17}.
The idea is to restrict the solution space of string variables to some pattern defined by \emph{flat automata} (see Section~\ref{sec:string-solving} for more details). 
The major benefit of considering this class is two folds.
First, even under the restriction, often the potential solution space is still infinite, which gives us a higher potential of finding solutions.
For instance, we can find a solution for $x.y \neq z  \wedge |x| > 2000$ under a very simple restriction: all variables are in $a^*$, which $a$ is a character in our alphabet.
Second, and more importantly, because we can convert the membership problem of a flat automaton to a satisfiability problem of Presburgher formula, the class of word equation + regular constraints + length constraints is decidable.

The paper of Abdulla et al. has considered adding string/number conversion constraints to the above class.
The paper proposes a decision procedure with a restricted form of flat automata.
For string/number conversion, their approach projects the solution to string/number conversion to, in terms of numerical solutions, a finite solution space.

@@

We propose various optimizations for satisfiability problem of quantifier-free $\paexp$ formula for an efficient implement. Two important optimizations are that we avoid the elimination of linear occurrences of variables by under approximation and synchronize the elimination of exponential occurrences of the same variable in different atomic formulas. We also apply an over approximation to reduce the number of subcases and adopt DFS with preprocessing to find possible solutions.

In fact, other than the theoretical difficulties, in practice, the string-number constraints are quite challenging for state-of-the-art solvers. 
Here we illustrate a toy example that mimic the ``mining'' step of block-chain construction. Essentially, giving a hash function ${\sf hash}(w)$, the goal of the mining step is to find an input $w$ such that ${\sf hash}(w)$ satisfies certain pattern, e.g., the last $k$ digits are zeros. Below we consider a simple hash function:
$${\sf hash}(w) = \sum \limits_{i = 1}^n a_i p^{n-i} \bmod m,$$ 
where $p, m \in \Int^+$ with $p, m \ge 2$. It is easy to see that ${\sf hash}(w)$ can be seen as a generalization of $\parseInt$ followed by a modulo operation. In particular, if $\Sigma = \Sigma_{\sf num}$ and $p = 10$, then ${\sf hash}(w) = \parseInt(w) \bmod m$. Thus the problem of finding a suitable input $w$ such that the last $k$ digits of ${\sf hash}(w)$ are zeros can be modeled as a $\strint$ formula. Although the example is seemingly simple, most state-of-the-art solvers have difficulty to handle it, c.f., Section~\ref{sec-exp} for our experimental results.

With several optimizations introduces in Section~\ref{sec-opt}, we manage to solve several variants of the hash function examples as well as some more randomly generated cases better than most state-of-the-art solvers.

\yfc{todo: a short description of optimizations}

>>>>

However, research progress of string constraint solving has been hampered by many major obstacles in both theory and tool implementation aspects (including long-standing open problems). 
Logical theories over strings have to allow string concatenation, which is arguably the most fundamental operation of strings. 
The most celebrated result concerning theories of strings is Makanin’s result on deciding the satisfiability problem for \emph{word equations}~\cite{makanin77}.
A simple example of a word equation is $xaby = ybax$, where $x, y$ are variables, and $a, b \in \Sigma$ are constant letters. 
A word equation is satisfiable if it has a solution, i.e., an assignment that maps variables to strings over the alphabet $\Sigma$ which equates the left-hand side with the right-hand side of the equation.
The correctness proof of Makanin’s algorithm is arguably one of the most complex termination proofs in computer science. 
Makanin’s result can be extended to include \emph{regular constraints} (a.k.a. regular expression matching, e.g., $x \in (ba)^*)$, and arbitrary Boolean connectives.
This extension is called word equations with regular constraints. 
However, the satisfiability problem of word equations together with length constraints (e.g., $|x|=|y|+1 \wedge wx=yx$) is still open.
The complexity of the satisfiability of word equations with regular constraints was proven to be PSPACE-complete by Plandowski~\cite{plandowski99}, after decades of improvement of the original algorithm by Makanin.

@@

The satisfiability of \emph{string/number conversion constraints}, e.g., $wx=yx \wedge |x| > \parseInt(y)$, has been proven undecidable in~\cite{DayGHMN18}. 
However, this kind of constraint is pervasive in scripting language programs. 
For example, it is common that programs read string input from text files and converts a part of the string input to numbers.
Even more crucially, in many programming languages, the string-number conversion is a part of the definition of their core semantics. 
JavaScript, which powers most interactive content on the Web and increasingly server-side code with Node.js, is one of such languages.

@@

One more recent approach is flattening~\cite{Parosh:20:PLDI,AbdullaACDHRR18,AbdullaACDHRR17}.
The idea is to restrict the solution space of string variables to some pattern defined by \emph{flat automata} (see Section~\ref{sec:string-solving} for more details). 
The major benefit of considering this class is two folds.
First, even under the restriction, often the potential solution space is still infinite, which gives us a higher potential of finding solutions.
For instance, we can find a solution for $x.y \neq z  \wedge |x| > 2000$ under a very simple restriction: all variables are in $a^*$, which $a$ is a character in our alphabet.
Second, and more importantly, because we can convert the membership problem of a flat automaton to a satisfiability problem of Presburgher formula, the class of word equation + regular constraints + length constraints is decidable.

The paper of Abdulla et al.~\cite{Parosh:20:PLDI} has considered adding string/number conversion constraints to the above class.
The paper proposes a decision procedure with a restricted form of flat automata.
For string/number conversion, their approach projects the solution to string/number conversion to, in terms of numerical solutions, a finite solution space.

@@

We propose various optimizations for satisfiability problem of quantifier-free $\paexp$ formula for an efficient implementation. Two important optimizations are that we avoid the elimination of linear occurrences of variables by under approximation and synchronize the elimination of exponential occurrences of the same variable in different atomic formulas. We also apply an over approximation to reduce the number of subcases and adopt DFS with preprocessing to find possible solutions.

In fact, other than the theoretical difficulties, in practice, the string-number constraints are quite challenging for state-of-the-art solvers. 
Here we illustrate a toy example that mimic the ``mining'' step of block-chain construction. 
Essentially, giving a hash function ${\sf hash}(w)$, the goal of the mining step is to find an input $w$ such that ${\sf hash}(w)$ satisfies certain pattern, e.g., the last $k$ digits are zeros. 
Below we consider a simple hash function:
$${\sf hash}(w) = \sum \limits_{i = 1}^n a_i p^{n-i} \bmod m,$$ 
where $p, m \in \Int^+$ with $p, m \ge 2$. 
It is easy to see that ${\sf hash}(w)$ can be seen as a generalization of $\parseInt$ followed by a modulo operation. 
In particular, if $\Sigma = \Sigma_{\sf num}$ and $p = 10$, then ${\sf hash}(w) = \parseInt(w) \bmod m$. Thus the problem of finding a suitable input $w$ such that the last $k$ digits of ${\sf hash}(w)$ are zeros can be modeled as a $\strint$ formula. Although the example is seemingly simple, most state-of-the-art solvers have difficulty to handle it, c.f., Section~\ref{sec-exp} for our experimental results. 
With several optimizations introduces in Section~\ref{sec-opt}, we manage to solve several variants of the hash function examples as well as some more randomly generated cases better than most state-of-the-art solvers.
